Assessing United States county-level exposure for tropical cyclone impact research

### 2020-07-10 ehp_revision/response_to_reviewers.Rmd

---
title: "Response to reviewers' comments for *\"Assessing United States county-level exposure for research on tropical cyclones and human health\"*"
output: bookdown::pdf_document2
toc: false
bibliography: writing/hurr_exposure.bib
header-includes:
- \usepackage{xcolor}
- \usepackage{framed}
- \usepackage{bm}
- \usepackage{subcaption}
- \usepackage{gensymb}
- \renewcommand\familydefault{\sfdefault}
- \usepackage{sansmath}
- \usepackage{bm}
- \usepackage{soul}
- \sansmath
- \usepackage{booktabs}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{pdflscape}
- \usepackage{float}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{fullpage}
- \usepackage{pdflscape}
- \usepackage{tcolorbox}
- \tcbuselibrary{skins,breakable}
- \usepackage{amsmath}
- \usepackage{xtab}
- \usepackage{rotating}
- \allowdisplaybreaks
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overall comments {-#overall-comments}

\colorlet{shadecolor}{blue!8}
\begin{shaded}
\textbf{Associate Editor's Overview:}
Developing a consistent and comparable metric for exposure to tropical cyclones
would be a valuable addition to the literature. All reviewers had questions
about the exposure metric, including the challenges of using county-level data
where not everyone in a county would have been exposed and the difficulties of
going from country-level to individual -level risks. Further the comparison of
different storm hazards could be clarified. The reviewers identified several
issues where further explanation would increase the accessibility of the
manuscript.
\end{shaded}

Thank you for the opportunity to revise and resubmit this manuscript. We
appreciate the helpful suggestions from the reviewers for clarifying and
deepening the discussion in the paper.

Broadly, we have:

1. Added discussion on the question of county-level versus individual-level
data. 
2. Added explanations for how our comparisons among different storm hazards
can help epidemiologists in designing studies and statistical analysis.
3. Added a small analysis of the correlation in rainfall metrics specifically
in cases of high rainfall ($\ge75$ mm of cumulative rainfall in the county
associated with the storm), to deepen the discussion of potential disagreement
in continuous measures of rainfall during extreme conditions. 
4. Changed labeling, axis range, and other presentation details for some
figures, as well as confirmed that the color scales will be accessible.
5. Corrected typos and improved wording throughout based on suggestions of the
reviewers.

@@

**Response:**

\begin{shaded}
\textbf{Reviewer 1 (R1) Comment 1:}
I very much appreciate the rationale for authors' decision to focus at the
county level, especially given the available health-related data. But, since
tropical cyclones don't necessarily have consistent impacts across an entire
county, I think the limitations of this decision are worthy of further
discussion in the discussion section.
\end{shaded}

**Response:**

- Ecological fallacy vs individual fallacy, especially in the disaster context
- Spatial misalignment / ecological effects of aggregating data at the county
(or other geopolitical) level. 
- Could be particularly problematic in some high-risk areas, because of the 
"long and narrow" characteristics of island counties. Dare County, Key West.

We think this is a very interesting question for disaster epidemiology, as
while there remains potential for ecological bias, the potential pathways for
health effects may also pass through damage in the individual's community beyond
their residence. For example, damage to major roads in the community might make
it harder for many residents to reach medical care in the aftermath of the
storm, even if their home did not experience extensive destruction. Similarly,
since power is provided over a network, conditions in an individual's community
can be important in determining risk of long-term outages (does Seth have
something on this?). As another example, if hospitals in the community are 
over capacity or have to evacuate, this could increase health risk for people in 
a fairly large "catchment" area for that hospital.

Therefore, while we agree that individual-level data would in many cases be 
helpful for tropical cyclone epidemiology, the ideal might be to design studies 
that integrate both community-level and individual-level metrics of exposure. 
This could help in determining if there are risks introduced to individuals based
on storm-related hazards in their broader community, after controlling for the 
hazards experienced directly at their homes.

@@

\begin{shaded}
\textbf{R1 Comment 2:}
Are there opportunities to expand this approach internationally? Some decisions
(e.g., county-level) data may restrict cross-country comparisons. Can the
authors discuss this? 
\end{shaded}

**Response:**

- Differs by exposure metric. Easier for some (wind, rain) than for others (flood, 
tornado)
- For rain, we use a re-analysis product only available for the continental US (right?).
However, there are similar gridded re-analysis products available for other regions 
worldwide that could be used in a similar way.
- For wind, the wind model could be generalizable with some tweaks. These include: 
  + More landmasks
  + Direction of storm movement for Northern vs Southern Hemispheres
  + Different wind averaging periods for the recorded best tracks data in different 
  storm basins.

@@

\begin{shaded}
\textbf{R1 Comment 3:}
In non-health fields, a county's inclusion in a FEMA major disaster declaration
or eligibility for public/individual assistance (all of which are somewhat
reflective of level of damage) is sometimes used as an "exposure" proxy. Have
you considered if/how your exposure assessments relate to inclusion in a FEMA
disaster declaration? In other words, does exposure to any of the hazards align
with the level of damage required for a county to be considered a "disaster
area" and/or worthy of federal assistance by current federal policy? Does this
matter?
\end{shaded}

**Response:**

This is a great question, and we are aware of a number of studies on the
societal impacts of tropical cyclones that have used FEMA disaster declarations
to assess county-level exposure to tropical cyclones [refs].

However, evidence from prior tropical cyclone epidemiology that doing so is
problematic and should be avoided [@grabich2015measuring]. FEMA's disaster
declarations are issued to provide assistance, based on damage assessment, to
individuals or the entire public in certain locations where catastrophes
overwhelm the state or local government [@mccarthy2014fema], and any use of them
for exposure assessment is secondary. Due to the political nature of disaster
declaration process, these declarations are often subject to many political and
economic factors [@mccarthy2014fema; @logue1981research]. One epidemiological
study specifically focused on exposure assessment for tropical cyclone
epidemiology and compared use of FEMA declarations with other methods of
exposure assessment, including one based directly on the storm hazard of wind
[@grabich2015measuring]. They found that exposure assessment based on FEMA
declarations tended to overassign counties to be "exposed," resulting in false
positives [@grabich2015measuring]. A further concern is that, given the
political and economic nature of these declarations, there is likely variation
across time and geography in the likelihood of a given exposure resulting in a
disaster declaration. This is particularly worrisome for a study that seeks to
explore risk across multiple years and affected communities.

We have added some discussion of this point...

- Future research could expand the work by [@grabich2015measuring] to determine
the extent of these issues with use of FEMA disaster declarations for
epidemiological exposure assessment. This would need to go beyond the comparison
with exposure assessment based on storm hazards, because sometimes the
characteristics of the affected community modifies the amount of disaster that
results from hazards of a certain severity. For example, a community that it
exposed often might be more "hardened" against damage from the storm (any work
from Seth on this) and so might experience less infrastructure damage from a
tropical cyclone with certain characteristics compared to a community that is
rarely exposed to tropical cyclones. One interesting direction would be to
investigate if there is evidence of differences in probablity of disaster
declarations at geopolitical boundaries, like when comparing counties on either
side of a state border.

@@

\textit{Lines 92--95 in the original manuscript:}

\begin{quotation}
\noindent
\textit{"In many cases, these studies analyze multi-year, multi-community data,
allowing them to estimate average associations over many disasters and to
explore how a disaster’s characteristics, or the characteristics of the
community it hits, modify associated health risks (e.g., Anderson and Bell 2010;
Son et al. 2012; Liu et al. 2017)."}
\end{quotation}
\end{shaded}

**Response:**

Thank you for this suggestion. We have changed the wording in this sentence:

> **Text in revised manuscript (relevant change in bold):**
"In many cases, these studies analyze multi-year, multi-community data, allowing
them to estimate average associations over many disasters and to explore how a
disaster’s characteristics, or the characteristics ofthe **affected
communities**, modify associated health risks (e.g., Anderson and Bell 2010; Son
et al. 2012; Liu et al. 2017)."

@@

\textit{Lines 137--141 from original manuscript:}

\begin{quotation}
\noindent
\textit{"While other approaches have been developed to incorporate storm
hazards, particularly wind, into exposure assessment (e.g., Grabich et al.
2015a; Zandbergen 2009; Czajkowski et al. 2011), there is not yet a standard
approach, and when different studies use different datasets or storm hazards
when assessing storm exposure, it becomes difficult to compare and aggregate
findings."}
\end{quotation}
\end{shaded}

**Response:**

This is a great suggestion. We have split the cited sentence into two sentences:

> **Text in revised manuscript:**
"While other approaches have been developed to incorporate storm
hazards, particularly wind, into exposure assessment (e.g., Grabich et al.
2015a; Zandbergen 2009; Czajkowski et al. 2011), there is not yet a standard
approach. When different studies use different datasets or storm hazards
when assessing storm exposure, it becomes difficult to compare and aggregate
findings."

@@

\begin{shaded}
\textbf{R1 Comment 6:}
This is not a comment on the paper, but the authors may also want to consider
publishing their work in DesignSafe, which is "the web-based research platform
of the NHERI  [NSF-Funded Natural Hazards Engineering Research Infrastructure]
Network that provides the computational tools needed to manage, analyze, and
understand critical data for natural hazards research." It is the data
repository widely used by interdisciplinary hazards and disaster researchers.
\end{shaded}

**Response:**

Great idea. We are investigating sharing this work in this repository.

@@

While I enjoyed reading the manuscript very much, I have one question I could
not figure out clearly.

One objective is to investigate patterns and agreement between different
exposure measures. I do not clearly understand the value of studying agreement
between exposures, while each exposure should be measured separately, so both
individual and overall effects can be assessed.  I do understand using "distance
from the storm" as exposure is only proxy measure and highly sensitive to
misclassification, but because direct storm hazards such as rainfall, flood,
storm surge, and tornado are publicly accessible as authors indicated, authors
could just simply guide readers to use the best exposure data rather than
conducting a quite intense analysis. This may be due to my limited knowledge, so
it would be appreciated if the authors provide explicit explanations on the
purpose and benefit of the analysis.

\end{shaded}

Thank you for this suggestion, and we can see that we could have more clearly 
articulated in the original manuscript how our assessment of patterns in and 
among exposures to specific hazards can help epidemiologists to design their
studies and the statistical methods to analyze the resulting data.

1. If exposure to the different hazards strongly agrees (i.e., when a county is 
exposed to one hazard, they are typically also exposed to a second), this would
have several implications for epidemiological research, especially for multi-year
studies: 
  + From a statistical modeling point of view, if you tried to include each 
  hazard as a separate independent variable, you would run into problems with 
  collinearity if fitting the study data with a generalized linear model or other
  regression model. One implication of this is that the confidence intervals on 
  effect estimates for the two variables would be very inflated, because the model
  would struggle to figure out which of the two variables the weight should be placed
  on, and so this uncertainty would be reflected in the size of the associated 
  confidence intervals.
  + If you see an effect, it would be very hard to try to determine if the causal 
  pathway goes through the first hazard or the second. To mitigate effects, it is 
  important to understand the pathway through which they happen, and it might be easier
  to get clues towards pathways if the hazards are better separated. (Analogy: if you
  were able to conduct a controlled experiment to look at the effects of each hazard, 
  jointly and in combination, you would design separation in the assignment of these
  treatments, so you could distinguish. We can't do that because we are limited to 
  an observational analysis, but the area of causal inference is doing a lot to 
  explore how we can design observational studies and analyze data to come closer to 
  what we might be able to see with a randomized control study, and which of those methods
  we might be able to use for tropical cyclone epidemiology will depend on the interplay of
  these patterns of exposures.)
2. These results will not be surprising to an atmospheric scientist / disaster researcher,
but there are environmental epidemiologists who have done research on other exposures and
are now starting to study tropical cyclones (session at ISEE, for example). It is useful 
for epidemiologists entering into tropical cyclone research to understand these facets of
tropical cyclone exposure, including the distinction between the very coastal wind exposures
and other hazards that tend to reach further inland. 
3. There are epidemiological studies that have used only one hazard for exposure assessment.
This has most often been the case for wind. If this exposure assessment is meant to 
capture "tropical cyclone exposure" in general, with all the hazards that are related to 
tropical cyclones, then there is the chance for exposure misclassification if this single
hazard is used. 
4. In the past, it was a very complex issue to try to pull together data on 
all these hazards. We agree that now that we have provided a dataset with all of them, 
epidemiologists should be thinking about multi-hazard analyses. However, there still 
will be some barriers to that. [Recent work on multi-pollutant exposures and the complexities
associated with that.] If one of the hazards dominates in explaining associated health 
risk, then in some cases it might be reasonable to focus on that hazard, but in that 
case it is critical, based on our results, to remember that the exposure assessment is
not capturing all hazards of the storms.

We have moved other suggestions that you sent in "Track Changes" of the Word 
document for the original manuscript and address each in detail below.

@@

\textit{Line 354 from original manuscript:}

\begin{quotation}
\noindent
\textit{"Finally, we investigated how well exposure assessment agreed across
these metrics."}
\end{quotation}
\end{shaded}

**Response:**

As we note in our general response to your comments, we agree in general with your
instinct here. However, there are some subtle points to note.

First, we understand the potential appeal of creating a single exposure 
summary score based on a summation of exposures. This simplifies some of the 
statistical modeling, for example, and also provides a single exposure estimate to 
provide as the key study conclusion. However, in many areas of tropical cyclone epidemiology, 
it is unlikely that exposure to each type of hazard should be given equal weight. For 
example, an analysis of respiratory health in the months after the storm might hypothesize
a strong connection with flooding, and smaller connections with rainfall and winds (through
the pathway of property damage). Use of a pre-defined summation of exposure to each hazard
would mask some of the potential association.

While a summary statistic could be defined that adds a weight to each exposure, it would
often be unclear *a priori* what weight each hazard exposure should be given. A better 
method might be to use a regression framework, where each hazard exposure is included as
a separate explanatory variable, and the model uses the observed data to determine the 
appropriate weight for each hazard in terms of modifying risk to the outcome of interest. 
To explore the influence of multiple concurrent hazards, interaction terms could be added.

However, if our analysis had found that hazard exposures follow very similar patterns, in 
terms of the counties exposed during a storm, then this strong agreement would result in 
a number of problems in fitting and interpreting these types of models. First, strong
agreement in two or more of the explanatory variables in a regression model leads to 
collinearity, which among other implications often leads to a large inflation of confidence
intervals for the effect estimates of the correlated explanatory variables. Further, if 
the hazard exposures were in strong agreement, then it would be difficult or impossible to
disentagle the effects of the separate hazards, as a strong effect for one would be 
impossible to interpret with confidence, since the strong agreement in the data (and resulting
low co-variability in the variables in the data) could cause the model to mistakenly 
attribute the effect of one hazard to the other. Further, it may become impossible in this
case to estimate synergistic effects between the two, as an indicator of an interaction
between the two exposures would be almost collinear with the two main effects.

In our analysis, we find that this is usually not the case for the four tropical cyclone 
hazards that we consider. Instead, we find that for most storms, the hazards are fairly 
well-separated in terms of the counties they effect. There is a particularly strong pattern
where inland counties can be exposed to extreme rain and flooding, but not wind. This 
separation opens the door for some interesting study designs and statistical modeling. 
For example, these patterns suggest that a study could investigate whether storm-associated
rains alone create a health risk, as there are many exposures with only rain in inland 
counties, and so health associations could be explored here and contrasted with coastal 
areas that experienced both wind and extreme rain.

@@

\textit{Line 392--394 from original manuscript:}

\begin{quotation}
\noindent
\textit{"Within these counties, storm-related rainfall measurements were
well-correlated between the two data sources, with rank correlations (bottom
right of each graph in Figure 2) between 0.87 and 0.98. "}
\end{quotation}
\end{shaded}

**Response:**

This is a great idea. In the original manuscript, we describe this lower correlation
at higher precipitation values through the use of examples from specific storms, and
discuss the potential reasons for it, but measuring the correlation specifically within
these higher-precipitation events would add more quantitative evidence to this discussion.

Here is the text where we discuss this point in the original manuscript:

> **Text in original manuscript:**
"There was some evidence that our primary rainfall metric may tend to
underestimate rainfall totals in storms with extremely high rainfall, based on
a few heavy-rainfall storms in Harris County, TX, Mobile County, AL, Charleston
County, SC, and Wake County, NC (Figure 2)."

> **Text in original manuscript:**
"The rainfall data are generally well-correlated with ground-based observations,
but may sometimes underestimate very high rainfall values (Figure 2). When
rainfall data is used to create binary exposure classifications, this
disagreement is unlikely to influence results, as both data sources agree in
identifying these as storms with high rainfall, but would be important to
consider for cases that include rainfall as a continuous measurement."

[Why we think this is happening... Re-analysis product uses a model to integrate
data from multiple sources and so might be likely to pull values back toward 
the "typical", and so oversmooth very extreme values.]

Here is the text and analysis we have added on this point in the revised manuscript:

@@

\textit{Line 491--495 from original manuscript:}

\begin{quotation}
\noindent
\textit{"For another set of tropical cyclones (e.g., Ernesto in 2006, and Bertha
in 1996, Isabel in 2003), there was moderate to good agreement for pairwise
combinations of distance, rain, and wind, but poor agreement for other
combinations of metrics, while for another set of storms (e.g., Matthew in 2004
and Katrina in 2005), there was moderate to good agreement between distance and
rain."}
\end{quotation}
\end{shaded}

**Response:**

[Check: *exposure matrix* in occupational epidemiology...]

[New advances in multi-pollutant analysis.]

[Some of the hazards could create similar risks, if they converge on a common
pathway of risk. For example, if pathway is through stress from property damage,
individually and in the county, then several hazards could indeed lead to the
same health outcomes (e.g., PTSD, other psychological markers, risk of acute
cardiovascular outcomes, etc.). One example is extreme level of destruction both
from wind (Hurricane Andrew, Hurricane Hugo), from river flooding associated
with among of precipitation (Hurricane Floyd, Hurricane Florence), and from
flash flooding associated with the rate of precipitation (Virginia storms).]

### 2020-07-10 ehp_revision/response_to_reviewers.Rmd

[A better future direction might be through regression models with multiple
explanatory pathways. This allows the model to fit different weights to each. Other
models can also be explored, like regression trees, which might have an easier time
capturing non-linearity (if exposure is included as a continuous value, as with 
wind and rain) and interactions when there are co-exposures to two or more hazards.
Both of these are able (and will) either exclude or give a very low weight to an 
explanatory variable that does not help much in explaining variation in the health 
outcome.]

@@

**Response:**

\begin{shaded}
\textbf{Reviewer 3 (R3) Comment 1:}
Starting on line 128, the authors write "Extreme winds are more common to the
track's right, where counterclockwise cyclonic winds move in concert with the
tropical cyclone's forward motion". I think this statement may be specific to
tropical cyclones in the Northern hemisphere (i.e., those relevant to exposure
in the US). Would be good to clarify this.
\end{shaded}

**Response:**

You are right, the opposite would be try in the Southern hemisphere. This
difference is driven by the Coriolis effect on cyclonic rotation, in which the
rotation of the earth diverts winds as they move toward the low pressure center
of the cyclone. We have edited the text to specify that this statement is
specific to the Northern Hemisphere:

> **Text in revised manuscript (relevant addition in bold):**
"**In the Northern Hemisphere,** extreme winds are more common to the track's
right, where counterclockwise cyclonic winds move in concert with the tropical
cyclone's forward motion."

@@

\begin{shaded}
\textbf{R3 Comment 2:}
Line 186: What is meant by the phrase "synoptic times"?

\textit{Lines 185--187 from original manuscript:}

\begin{quotation}
\noindent
\textit{"We used tracking data from HURDAT2, which records the storm center’s
position at the synoptic times of 6:00 am, 12:00 pm, 6:00 pm, and 12:00 am
Coordinated Universal Time (UTC)."}
\end{quotation}
\end{shaded}

@@

\begin{shaded}
\textbf{R2 Comment 3:}
Several of the exposures are calculated for the population mean center of each
county. From the reference list, it looks like the authors consistently used
population centers from 2010 throughout the years included in the dataset/R
package. If so, I suggest clarifying this in the text. 
\end{shaded}

**Response:**

\begin{shaded}
\textbf{R2 Comment 4:}
In the methods section starting on line 276, the authors describe the flooding
and tornado data from the NOAA storm events database. I'd suggest providing some
information about where the events in this database come from (e.g., that they
are reported events) here in addition to covering it in the discussion.

\textit{Lines 276--281 from original manuscript:}

\begin{quotation}
\noindent
\textit{"To identify flood- and tornado-based tropical cyclone exposures in US
counties, we matched storm tracks with event listings from the National Oceanic
and Atmospheric Administration (NOAA)’s Storm Event Database (NOAA NCEI 2020).
While this database has recorded storm data, particularly tornadoes, since 1950,
its coverage changed substantially in 1996 to cover more types of storm events,
including flood events (NOAA NCEI 2020). We therefore only considered flood
metrics of tropical cyclone exposure for storms in 1996 and later."}
\end{quotation}
\end{shaded}

@@

**Response:**

\begin{shaded}
\textbf{R2 Comment 5:}
Line 294: What is meant by the phrase "traditional tornado event database"?

\textit{Lines 294--295 from original manuscript:}

\begin{quotation}
\noindent
\textit{"The tornado observations from this dataset form a traditional tornado
event database for the US, and so we did not further validate the tornado event
data."}
\end{quotation}
\end{shaded}
